---
title: "Untitled"
output:
  html_document:
    toc: true
    fig_caption: true
---

## Synopsis

## Data Processing

```{r cache=TRUE}

# checking if data directory exists, if not, creating one
if(!file.exists("./data")) { dir.create("./data") }

# location of data to download
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# name of file
fileName <- "data/repdata-data-StormData.csv.bz2"

# downloading the file and placing it in the data directory if it doesn't already exist
if(!file.exists(fileName) { download.file(fileURL, fileName, method="curl")}

# reading the data into R
data <- read.csv(fileName)

# let's use dplyr
library(dplyr)
data <- tbl_df(data)

```

## Results

### Which Events are Most Harmful to Population Health?

To answer this question we will look at Event Types that result in direct fatalities or injuries. Because many of the event types are inconsistently recorded (ex. THUNDERSTORM WIND and THUNDERSTORM WINDS) I'm going to use two processes to clean up the data:

* converting all events to lowercase, which removes capitalization errors
* approximate string matching using the generalized Levenshtein distance.

```{r}
# data frame with event type, fatalities and injuries
harmfulevents <- data %>% 
  select(EVTYPE, FATALITIES, INJURIES) %>%
  # all lower case to clean up some event types
  mutate(EVTYPE = tolower(EVTYPE)) %>%
  group_by(EVTYPE) %>%
  summarize(total = sum(FATALITIES,INJURIES)) %>%
  arrange(desc(total))

```



```{r}

combineevents <- function(df) {
}



```

### Which Events have the Greatest Economic Consequence?

